---
title: "Homework 1"
author: "Triveni Sangama Saraswathi"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

CS 625, Fall 2020

## Git, GitHub

1. *What is your GitHub username?*
  
    Triveniedla

2. *What is the URL of your remote GitHub repo (created through Mr. Kennedy's exercises)?*

     git@github.com:Triveniedla/cs625-fall-2020.git

## R

The command below will load the tidyverse package.  If you have installed R, RStudio, and the tidyverse package, it should display a list of loaded packages and their versions.
```{r}
library(tidyverse)
```

## R Markdown 

1. *Create an ordered bulleted list with at least 3 items*

*   Bulleted list item 1

*   Item 2

    * Item 2a

    * Item 2b

2. *Write a paragraph that demonstrates the use of italics, bold, bold italics, and code.*

3. *Create an example of a fenced code block.*

4. *Create a level 4 heading*

## R 

#### Data Visualization Exercises

1. *Run ggplot(data = mpg). What do you see?*
   
   ```{r}
'ggplot(data = mpg)'
```
   
   R is not showing any output.

2. *How many rows are in mpg? How many columns?*
   
   There are 234 rows and 11 columns.

3. *What does the drv variable describe? Read the help for ?mpg to find out.*
   
   It describes the type of drive train, where f=front-wheel drive, r=rear wheel drive, 4=4wd.
  

4. *Make a scatterplot of hwy vs cyl.*


   ```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cyl, y = hwy))

```

5. *What happens if you make a scatterplot of class vs drv? Why is the plot not useful?*

 ```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = drv))

```
The attributes class and drv are not showing any relationship.

#### Workflow: basics Exercises

1. *Why does this code not work?*

```r
my_variable <- 10
my_varıable
```

2. *Tweak each of the following R commands so that they run correctly:*

```r
library(tidyverse)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

fliter(mpg, cyl = 8)
filter(diamond, carat > 3)
```

3. *Press Alt + Shift + K. What happens? How can you get to the same place using the menus?*

## Tableau

*Insert your the image of your final bar chart here*

1. *What conclusions can you draw from the chart?*


## Observable and Vega-Lite

### A Taste of Observable

1. *In the "New York City weather forecast" section, try replacing `Forecast: detailedForecast` with `Forecast: shortForecast`. Then press the blue play button  or use Shift-Return to run your change. What happens?*

1. *Under the scatterplot of temperature vs. name, try replacing `markCircle()` with `markSquare()`. Then press the blue play button  or use Shift-Return to run your change. What happens? How about `markPoint()`?*

1. *Under "Pick a location, see the weather forecast", pick a location on the map.  Where was the point you picked near?*

1. *The last visualization on this page is a "fancy" weather chart embedded from another notebook.  Click on the 3 dots next to that chart and choose 'Download PNG'.  Insert the PNG into your report.*

### Charting with Vega-Lite

1. *Pass an option of { size: 200 } to markCircle().*

1. *Try markSquare instead of markCircle.*

1. *Try markPoint({ shape: 'diamond' }).*

1. *Change Horsepower to Acceleration*

1. *Swap what fields are displayed on the x- and y-axis* 

1. *Change Name to Origin.*

1. *Remove the vl.y().fieldN("Origin") line.*

1. *Replace count() with average("Miles_per_Gallon").*

## References

*Insert the list of sites you used as references as an unordered list with named links here.  This is required.*
