---
title: "Homework 5"
author: "Triveni Sangama Saraswathi Edla"
output: rmarkdown::github_document
---


CS 625, Fall 2020

In this homework, R is used to generate the __boxplot__, __eCDF__ and __histograms__ for different datasets.

### 1. [NYT Mask Usage Survey dataset](https://github.com/nytimes/covid-19-data/tree/master/mask-use):


In this section, the NYT mask survey data is analyzed by plotting distributions. The dataset contains the share of people answered for the mask survey question as __NEVER__, __RARELY__, __SOMETIMES__, __FREQUENTLY__, and __ALWAYS__. First, the __boxplot__is plotted for distributions for different answers.    


```{r}
library(ggplot2)
library(tidyverse)

mask_data <- read.csv(file='mask-use-by-county.csv')

NEVER = data.frame(answers = "NEVER", value = mask_data$NEVER)
RARELY = data.frame(answers = "RARELY", value =mask_data$RARELY)
SOMETIMES = data.frame(answers = "SOMETIMES", value =mask_data$SOMETIMES)
FREQUENTLY = data.frame(answers = "FREQUENTLY", value =mask_data$FREQUENTLY)
ALWAYS = data.frame(answers = "ALWAYS", value =mask_data$ALWAYS)

plot.data = rbind(NEVER ,RARELY,SOMETIMES,FREQUENTLY,ALWAYS)

p<- ggplot(plot.data, aes(x=answers, y=value,fill=answers))+ geom_boxplot()

p<-p+labs(x = "Mask survey answers",y = "Share of people in each county",face =c("plain"))

p<-p+theme_bw()+theme(legend.position = "none")

p
```


For plotting the __boxplot__, dataframes are created for each answer and the dataframes are binded using _rbind_ feature of R. From the plot, it can be observed that the median of the share of people, of different counties, who answered __ALWAYS__ are higher followed by __FREQUENTLY__, __SOMETIMES__, __RARELY__ and __NEVER__. Also, outliers are not observed in the share of people who answered __ALWAYS__.

The __advantages__ of using boxplot is it handles the large data easily and displays the outliers in the data along with a clear summary. The __disadvantages__ of using boxplot is that it hides the multimodality and other features of distributions. It confuses some audiences and also difficult to locate sometimes. I find boxplot is suitable for this type of dataset as it is easy to compare the features from the plot.


Next, the empirical cumulative distribution function (__eCDF__) plot is plotted using _stat_ecdf_ feature for a share of people answered __ALWAYS__ or __FREQUENTLY__ as shown below.



```{r}

library(ggplot2)

mask_data$combined=mask_data$FREQUENTLY+mask_data$ALWAYS

p<-ggplot(mask_data, aes(combined)) + stat_ecdf(geom = "step")

p<-p+labs(x = "Frequently/Always",y = "CDF",face =c("plain"))

p<-p+theme_bw()

p


```

The above plot shows that the cumulative distribution line is increasing and share of people who  answered either __FREQUENTLY/ALWAYAS__ are mostly between 0.4 and 1.0 for different counties. However, it is hard to predict or compare the features in the dataset. The ECDF is an estimator that allows us to plot features of our data in order from least to greatest. I observe that ECDF is not the best plot for the NYT MASK DATA as it is hard to analyze from the plot.

Below are the histograms generated for the share of people answered __ALWAYS__ or __FREQUENTLY__ with two different bin sizes using the _geom_histogram()_ feature of R. Further, the default labels are replaced by custom labels using the _labs_ feature, and the background of the plot is changed using _theme_bw()_ feature.

```{r}

library(ggplot2)

p<-ggplot(mask_data, aes(x=combined))+geom_histogram(binwidth=0.01,color="black", fill="blue")

p<-p+labs(x = "Frequently/Always",y = "count",face =c("plain"))
p<-p+theme_bw()
p

p<-ggplot(mask_data, aes(x=combined))+geom_histogram(binwidth=0.03,color="black", fill="blue")

p<-p+labs(x = "Frequently/Always",y = "count",face =c("plain"))
p<-p+theme_bw()
p
```

From the plot, it can be observed that the mode of share of people, for all the counties, who have answered __FREQUENTLY__ or __RARELY__ is around 80%. Further, the histogram is skewed normal distribution towards left, with mean shifted to the left. The histogram with higher width tends towards a smother distribution and however much higher width does not give clear information about the distribution.


The __advantages__ of the histogram is simplicity and versatility to work with large ranges of information in the data. It is used in many situations to offer an insightful look at frequency distribution. The __disadvantages__ of the histogram are the observations that could be misleading sometimes because of too many blocks. 

From the three plots, I feel that _boxplot_ is better for the NYT MASK DATA than _eCDF_ and _histogram_.



## Part 2: [US Census Bureau County Population dataset](https://www.census.gov/data/datasets/time-series/demo/popest/2010s-counties-total.html#par_textimage_70769902):

In part2, the US Census dataset is analyzed. The data has several columns in which the __CENSUS2010POPULATION__ column has been selected and plotted using _boxplot_. First, the plot is generated using the original scale as shown below.


```{r}
library(ggplot2)
library(tidyverse)
library(scales)

census_bureau <- read.csv(file='co-est2019-alldata.csv')


CENSUS2010POP = data.frame(answers = "CENSUS2010POP", value = census_bureau$CENSUS2010POP)

plot.data = rbind(CENSUS2010POP)

p<-ggplot(plot.data, aes(x=answers, y=value, fill=answers)) +  geom_boxplot()

p<-p+labs(x = "",y = "Population",face =c("plain"))
p<-p+theme_bw()+ theme(legend.position = "none")
p

```

The _boxplot_ shows a lot of outliers and difficult to observed the statistics from the plot. To overcome this problem, a log scale is used to plot the data as shown below. For this purpose, _scales_ package is used and _log10_trains()_ feature is used in the _scale_y_continuous()_ function.

```{r}
library(ggplot2)
library(tidyverse)
library(scales)

census_bureau <- read.csv(file='co-est2019-alldata.csv')


CENSUS2010POP = data.frame(answers = "CENSUS2010POP", value = census_bureau$CENSUS2010POP)

plot.data = rbind(CENSUS2010POP)

p<-ggplot(plot.data, aes(x=answers, y=value, fill=answers)) +  geom_boxplot()

p<-p+labs(x = "",y = "Population",face =c("plain"))
p<-p+theme_bw()+ theme(legend.position = "none")

p <- p + scale_y_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) 

p
```

The _boxplot_ on log scale shows very good information with median and Q1, Q2 quantiles for the states pouplations is around 4x $$10^4$$, $$10^4$$ and 9x $$10^4$$ respectively. To improve the visualization of boxplot further, 25%, 50%, 75%, 90%, 95%, and 99% quantiles is extracted from the data as shown below.

```{r}
prcntl_values <- quantile(CENSUS2010POP$value, c(.25, .50, .75, .90, .95, .97, .99))
prcntl_values
```
The data points under 99% quantiles are with a population below 3588277. The data points with a population above this value are removed and box plots are generated using normal and log scales as shown below.



```{r}
CS2010_99 <- CENSUS2010POP[with(CENSUS2010POP, value > prcntl_values[7]),]
plot.data = rbind(CS2010_99)
p<-ggplot(plot.data, aes(x=answers, y=value, fill=answers)) +  geom_boxplot()
p<-p+labs(x = "",y = "Population",face =c("plain"))
p<-p+theme_bw()+ theme(legend.position = "none")
p
p <- p + scale_y_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) 
p
```

The _boxplots_ shows good information with both normal and log scales but some of the data is removed in the above plots. The _boxplot_ with log scale, including the data points above 99 percentile, also shows good information. Hence, the entire data is used for _eCDF_, _histogram_ plots, and normal scale and log scales plots are compared.

```{r}

library(ggplot2)
library(scales)

p<-ggplot(CENSUS2010POP, aes(value)) + stat_ecdf(geom = "step")

p<-p+labs(x = "CENSUS2010POP",y = "CDF",face =c("plain"))
p<-p+theme_bw()
p

p <- p + scale_x_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) 

p

```

The _eCDF_ plot on the normal scale does not give clear information about the distribution and whereas, the plot on the log scale shows good information. From _eCDF_, it can be observed that most of the people are between  $$10^3$$ and  $$10^7$$ and it does not show information about the outlines.

Next, histograms are plotted for __CENSUS2010POP__ data of all the counties with two different bin widths only in log scales as the plots in the normal scale are not showing good information.

```{r}

p<-ggplot(CENSUS2010POP, aes(x=value))+geom_histogram(binwidth=0.1,color="black", fill="blue")
p<-p+labs(x = "Frequently/Always (log10 scale)",y = "count",face =c("plain"))
p<-p+theme_bw()

p <- p + scale_x_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) 

p

p<-ggplot(CENSUS2010POP, aes(x=value))+geom_histogram(binwidth=0.2,color="black", fill="blue")
p<-p+labs(x = "Frequently/Always (log10 scale)",y = "count",face =c("plain"))
p<-p+theme_bw()

p <- p + scale_x_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) 

p

```
From the histogram, it is observed that the distribution is close to a left tailed normal with the mode around 3e4.   

3. [US Census Bureau State Population dataset](https://www.census.gov/data/tables/time-series/demo/popest/2010s-state-total.html#par_textimage)

In this section, the population data of each state is visualized. The data set contains the population of each county of all states and the visualization need to be conducted for each state. For this purpose, the population of all counties in each state is combined. 

First, a dataframe ( __est2019req__ ) is created for state name( __STNAME__ ) ,birth rates ( __BIRTHS2019__ ), death rates( __DEATHS2019__ ) and migration rates( __RINTERNATIONALMIG2019__ ). From this dataframe ( __est2019req__ ), the population with same states are aggregated using __aggregate()__ function of R as shown below. 


```{r}
library(ggplot2)
library(tidyverse)


est2019 <- read.csv(file='co-est2019-alldata.csv')

est2019req <- data.frame(STNAME=est2019$STNAME,
                         BIRTHS2019=est2019$BIRTHS2019,
                         DEATHS2019=est2019$DEATHS2019,
                         RINTERNATIONALMIG2019=est2019$RINTERNATIONALMIG2019)

est2019req=aggregate(. ~ STNAME, data=est2019req, FUN=sum)
```

Next, the boxplot is generated for all the columns in the __est2019req__ dataframe.

```{r}

BIRTHS2019 = data.frame(answers = "BIRTHS2019", value = est2019$BIRTHS2019)
DEATHS2019 = data.frame(answers="DEATHS2019", value = est2019$DEATHS2019)
RINTERNATIONALMIG2019 = data.frame(answers="RINTERNATIONALMIG2019", value = est2019$RINTERNATIONALMIG2019)


plot.data = rbind(BIRTHS2019, DEATHS2019, RINTERNATIONALMIG2019)

p1<- ggplot(plot.data, aes(x=answers, y=value,fill=answers))+ geom_boxplot()
p1<-p1+labs(x = "Birth,death and migration rates",y = "Share of people in each state",face =c("plain"))
p1<-p1+theme_bw()+theme(legend.position = "none")
p1 <- p1 + scale_y_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x)))

p1


```

The __boxplot__ shows a lot of outliers for all the distributions. From the plot is observed that the median birth rates and death rates are almost the same and the median migration rates are lower than the other two.

Below are the __eCDF__ and histogram plots for birth rates ( __BIRTHS2019__ ).

```{r}
library(ggplot2)

p2<-ggplot(est2019req, aes(BIRTHS2019)) + stat_ecdf(geom = "step")

p2<-p2+labs(x = "est2019 ",y = "BIRTHS2019",face =c("plain"), inherit.aes = FALSE)

p2<-p2+theme_bw()
p2 <- p2 + scale_x_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x)))

p2
```

The variation of cumulative distribution function in the __eCDF__ plot is very rough. This is because there are only 50 data points as there are only 50 states in the united states. From the plot, it is observed that the population of the states are between $$10^4$$ and $$10^6$$.


```{r}
p<-ggplot(est2019req, aes(x=BIRTHS2019))+geom_histogram(binwidth=0.1,color="black", fill="blue")
p<-p+labs(x = "Frequently/Always (log10 scale)",y = "count",face =c("plain"))
p<-p+theme_bw()

p <- p + scale_x_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) 

p


```
```{r}
p<-ggplot(est2019req, aes(x=BIRTHS2019))+geom_histogram(binwidth=0.2,color="black", fill="blue")
p<-p+labs(x = "Frequently/Always (log10 scale)",y = "count",face =c("plain"))
p<-p+theme_bw()

p <- p + scale_x_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) 

p


```

The histograms for the birth rates for all states do not give good information because there are only a few data points. For the state population dataset, __boxplot__ shows good information compared to histogram and __eCDF__



## References:

* Creating dataframe in R: https://stackoverflow.com/questions/11346880/r-plot-multiple-box-plots-using-columns-from-data-frame
* Creating boxplot using ggplot: http://www.sthda.com/english/wiki/ggplot2-box-plot-quick-start-guide-r-software-and-data-visualization

* Creating ecdf using ggplot: http://www.sthda.com/english/wiki/ggplot2-ecdf-plot-quick-start-guide-for-empirical-cumulative-density-function-r-software-and-data-visualization

* Creating histogram using ggplot: http://www.sthda.com/english/wiki/ggplot2-histogram-plot-quick-start-guide-r-software-and-data-visualization

* Calculating percentile statistics in R: https://www.r-tutor.com/elementary-statistics/numerical-measures/percentile


