---
title: "Homework 8"
author: "Triveni Sangama Saraswathi Edla"
Due: "Friday, December 11, 2020 by 11:59pm"
output: rmarkdown::github_document
---

In the previous assignment, datasets of different business sectors affected by the COVID-19 pandemic was explored and questions were generated. For each question, a draft chart was proposed. In this homework, the draft chart is refined for the below question.

### How does the shortage of beef production rates during the COVID-19 pandemic affect the beef price?

Below is the code used from homework 8 and it cleans the data, that will be used for developing the refined chart. For designing the chart, the COVID-19 data, beef production rates, and beef prices are used. 

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(scales)

data_coronavirus <- read.csv(file='./cleaned_data_files/us_coronavirus.csv')
data_coronavirus$date <- ymd(data_coronavirus$date)
#creating new column with number of new cases
data_coronavirus <- within(data_coronavirus, new_cases <- c(0,diff(cases)))

data_beef<- read.csv(file = './cleaned_data_files/us_beef.csv')
data_beef <- data_beef%>% select(Year, Label, Value)
data_beef <- data_beef %>% separate(Label, c("y", "Month"))
data_beef <- data_beef%>% select(Year, Month, Value)
data_beef$Ground=data_beef$Value
data_beef <- data_beef%>% select(Year, Month, Ground)

data_dum<- read.csv(file = './cleaned_data_files/us_beef_chunk.csv')
data_beef$Chunk=data_dum$Value
data_dum<- read.csv(file = './cleaned_data_files/us_beef_uncooked.csv')
data_beef$Uncooked=data_dum$Value

#Converting month to number
data_beef$Month=match(data_beef$Month,month.abb)
#Combining year and month
data_beef$Date <- with(data_beef, sprintf("20-%02d-28", data_beef$Month))
#Converting to date date datatype
data_beef$Date<- ymd(data_beef$Date)
data_beef <- data_beef%>% select(Year,Date,Ground,Chunk,Uncooked)

# library(zoo)

data_meatproduction<- read.csv(file = './cleaned_data_files/us_beef_chicken_production.csv')
data_meatproduction<- data_meatproduction %>% separate(Date, c("Month", "Year"))

#Converting comma separated character to integer and calcaulting net beef and chicken
data_meatproduction$Beef=as.numeric(gsub(",", "", data_meatproduction$Beef))
data_meatproduction$NetBeef<-data_meatproduction$Beef+data_meatproduction$Veal
data_meatproduction$Broilers=as.numeric(gsub(",", "", data_meatproduction$Broilers))
data_meatproduction$Netchicken<-data_meatproduction$Broilers+data_meatproduction$Other_chicken


data_meatproduction$Year=as.numeric(gsub(",", "", data_meatproduction$Year))

#Converting month to number
data_meatproduction$Month=match(data_meatproduction$Month,month.abb)
#Combining year and month
data_meatproduction$Date <- with(data_meatproduction, sprintf("20-%02d-28", data_meatproduction$Month))
#Converting to date datatype
data_meatproduction$Date<- ymd(data_meatproduction$Date)
data_meatproduction <- data_meatproduction%>% select(Year,Date,NetBeef,Netchicken)

```


From the above code, the data frames ```data_coronavirus```, ```data_beef``` and ```data_meatproduction``` is used. 

#### Original draft chart

```{r}

library("ggpubr")

p1<-ggplot(data=data_coronavirus, aes(x=date, y=new_cases)) +  geom_line()
p1<-p1+theme(axis.text.x = element_text(angle=90, hjust = 1))
p1<-p1+scale_x_date(labels = date_format("%b"))
p1<-p1+labs(x = "",y = "New COVID cases")

p2<-ggplot()
p2<-p2+geom_line(data=data_beef %>% filter(Year==2020), aes(x=Date, y=Ground, color = "2020"),linetype="solid",lwd = 1)
p2<-p2+geom_line(data=data_beef %>% filter(Year==2019), aes(x=Date, y=Ground,color = "2019" ),linetype="dotted",lwd = 1)
p2<-p2+geom_line(data=data_beef %>% filter(Year==2018), aes(x=Date, y=Ground, color = "2018"),linetype="dashed",lwd = 1)
p2<-p2+scale_color_discrete(name = "Ground")
p2<-p2+theme(axis.text.x = element_text(angle=90, hjust = 1),legend.position = "none",legend.justification = c("right", "top"),plot.title = element_text(face = "plain",hjust = 0.5),legend.title=element_blank())
p2<-p2+scale_x_date(labels = date_format("%b"))
p2<-p2+labs(x = "Month",y = "Average price (per lb)",title = "Ground")

p3<-ggplot()
p3<-p3+geom_line(data=data_beef %>% filter(Year==2020), aes(x=Date, y=Chunk, color = "2020"),linetype="solid",lwd = 1)
p3<-p3+geom_line(data=data_beef %>% filter(Year==2019), aes(x=Date, y=Chunk,color = "2019" ),linetype="dotted",lwd = 1)
p3<-p3+geom_line(data=data_beef %>% filter(Year==2018), aes(x=Date, y=Chunk, color = "2018"),linetype="dashed",lwd = 1)
p3<-p3+scale_color_discrete(name = "Ground chunk")
p3<-p3+theme(axis.text.x = element_text(angle=90, hjust = 1),legend.position = "none",legend.justification = c("right", "top"),plot.title = element_text(face = "plain",hjust = 0.5))
p3<-p3+scale_x_date(labels = date_format("%b"))
p3<-p3+labs(x = "Month",y = "Average price (per lb)",title = "Chunk")

p4<-ggplot()
p4<-p4+geom_line(data=data_beef %>% filter(Year==2020), aes(x=Date, y=Uncooked, color = "2020"),linetype="solid",lwd = 1)
p4<-p4+geom_line(data=data_beef %>% filter(Year==2019), aes(x=Date, y=Uncooked,color = "2019" ),linetype="dotted",lwd = 1)
p4<-p4+geom_line(data=data_beef %>% filter(Year==2018), aes(x=Date, y=Uncooked, color = "2018"),linetype="dashed",lwd = 1)
p4<-p4+scale_color_discrete(name = "Uncooked ground")
p4<-p4+theme(axis.text.x = element_text(angle=90, hjust = 1),legend.position = "none",legend.justification = c("right", "top"),plot.title = element_text(face = "plain",hjust = 0.5))
p4<-p4+scale_x_date(labels = date_format("%b"))
p4<-p4+labs(x = "Month",y = "Average price (per lb)",title = "Uncooked")

p5<-ggplot()
p5<-p5+geom_line(data=data_meatproduction %>% filter(Year==2020), aes(x=Date, y=NetBeef, color = "2020"),linetype="solid",lwd = 1)
p5<-p5+geom_line(data=data_meatproduction %>% filter(Year==2019), aes(x=Date, y=NetBeef,color = "2019" ),linetype="dotted",lwd = 1)
p5<-p5+geom_line(data=data_meatproduction %>% filter(Year==2018), aes(x=Date, y=NetBeef, color = "2018"),linetype="dashed",lwd = 1)
p5<-p5+scale_color_discrete(name = "Beef")
p5<-p5+theme(axis.text.x = element_text(angle=90, hjust = 1),legend.position = "none",legend.justification = c("right", "top"),plot.title = element_text(face = "plain",hjust = 0.5))
p5<-p5+scale_x_date(labels = date_format("%b"))
p5<-p5+labs(x = "Month",y = "Production (M lbs)")

figure <- ggarrange(p1, p2,p3, p5, p4,ncol = 3, nrow = 2,  common.legend = TRUE)
figure

```

The above chart is the original draft chart that was proposed in homework 7 to answer the question mentioned at the start of this homework. It is refined in this assignment in the below section. 

#### Refined chart

```{r}

heading="COVID-19 Outbreak in Meat Plants Caused Shutdowns, Dropped Meat Production,\n and Increased Beef Product Prices"
heading2="Presidential Executive Order Resumed Meat Production and Drop-in Price of Beef Products"

p1<-ggplot()
p1<-ggplot(data=data_coronavirus, aes(x=date, y=new_cases/100000)) +  geom_line(color="blue")
p1<-p1+scale_x_date(labels = date_format("%b"),limits = as.Date(c("2020-01-27","2020-12-01")),date_breaks = "months")
p1<-p1+labs(y = "New cases",title=heading, subtitle = heading2)
p1<-p1+theme(legend.position = "none",plot.title = element_text( family="serif",face="bold", colour="blue", size = 13.0))
p1<-p1+theme(plot.subtitle = element_text( family="serif",face="italic", colour="black", size = 11.0))

p1<-p1+theme(axis.title.x = element_blank(),axis.ticks = element_blank(), panel.grid.minor.x = element_blank())
p1<-p1+font("ylab")

data2020=data_meatproduction %>% filter(Year==2020)
data2019=data_meatproduction %>% filter(Year==2019)
data2018=data_meatproduction %>% filter(Year==2018)
p2<-ggplot()
p2<-p2+geom_line(data=data2020, aes(x=Date, y=NetBeef/1e3, color = "2020"),linetype="solid",lwd = 1)
p2<-p2+geom_line(data=data2019, aes(x=Date, y=NetBeef/1e3,color = "2019" ),linetype="dotted",lwd = 1)
p2<-p2+geom_line(data=data2018, aes(x=Date, y=NetBeef/1e3, color = "2018"),linetype="dashed",lwd = 1)
p2<-p2+scale_x_date(labels = date_format("%b"),limits = as.Date(c("2020-01-27","2020-12-01")),date_breaks = "months")
p2<-p2+theme(legend.position = "none",legend.title=element_blank())
p2<-p2+labs(y = "Meat production")
p2<-p2+theme(axis.title.x = element_blank(),axis.ticks = element_blank(), panel.grid.minor.x = element_blank())
p2<-p2+font("ylab")

#Covid outbreak
p2<-p2+ geom_point(aes(x = as.Date(c("2020-03-15")), y = 2.26),size=5.0,color="orange")

#Production suspension
datesuspen=as.Date(c("2020-03-31",
                     "2020-04-02",
                     "2020-04-7",
                     "2020-04-8",
                     "2020-04-9",
                     "2020-04-10",
                     "2020-04-13",
                     "2020-04-14",
                     "2020-04-16",
                     "2020-04-17",
                     "2020-04-18",
                     "2020-04-20",
                     "2020-04-21",
                     "2020-04-22",
                     "2020-04-23",
                     "2020-04-24",
                     "2020-04-27"))
prodapprox=approx(data2020$Date,data2020$NetBeef, xout=datesuspen)$y
p2<-p2+ geom_point(aes(x = datesuspen, y = prodapprox/1e3),size=1.5,color="red")

prodexec=approx(data2020$Date,data2020$NetBeef, xout=as.Date(c("2020-04-29")))$y
p2<-p2+ geom_point(aes(x = as.Date(c("2020-04-29")), y = 1.01*prodexec/1e3),size=5.0,color="darkgreen")

#Production start
datestart=as.Date(c( "2020-05-04",
                     "2020-05-5",
                     "2020-05-6",
                     "2020-05-7",
                     "2020-05-8",
                     "2020-05-11",
                     "2020-05-13",
                     "2020-05-15",
                     "2020-05-19",
                     "2020-05-21",
                     "2020-05-29",
                     "2020-06-04",
                     "2020-06-09",
                     "2020-06-16"))
prodstart=approx(data2020$Date,data2020$NetBeef, xout=datestart)$y
p2<-p2+ geom_point(aes(x = datestart, y = prodstart/1e3),size=1.5,color="green3")

p2<-p2+ geom_text(aes(x = as.Date(c("2020-02-16")), y = 2.37,label = "Outbreak"), size=3.2)
p2<-p2+geom_segment(aes(x = as.Date(c("2020-03-03")), y = 2.33, xend = as.Date(c("2020-03-13")), yend = 2.275),
                  arrow = arrow(length = unit(0.3, "cm")),size=1.00)

p2<-p2+ geom_text(aes(x = as.Date(c("2020-05-09")), y = 2.38,label = "Shutdowns"), size=3.2)
p2<-p2+geom_segment(aes(x = as.Date(c("2020-04-20")), y = 2.38, xend = as.Date(c("2020-04-04")), yend = 2.33),
                  arrow = arrow(length = unit(0.3, "cm")),size=1)

p2<-p2+ geom_text(aes(x = as.Date(c("2020-03-22")), y = 1.82,label = "Executive order"), size=3.2)
p2<-p2+geom_segment(aes(x = as.Date(c("2020-04-16")), y = 1.80, xend = as.Date(c("2020-04-25")), yend = prodexec/1e3),
                  arrow = arrow(length = unit(0.3, "cm")),size=1)


p2<-p2+ geom_text(aes(x = as.Date(c("2020-07-28")), y = 2.0,label = "Resumed production"), size=3.2)
p2<-p2+geom_segment(aes(x = as.Date(c("2020-06-26")), y = 2.0, xend = as.Date(c("2020-06-12")), yend = 2.03),
                  arrow = arrow(length = unit(0.3, "cm")),size=1)


data_beef$average<-(data_beef$Ground+data_beef$Chunk+data_beef$Uncooked)/3
p3<-ggplot()
p3<-p3+geom_line(data=data_beef %>% filter(Year==2020), aes(x=Date, y=average,color = "2020"),linetype="solid",lwd = 1)
p3<-p3+geom_line(data=data_beef %>% filter(Year==2019), aes(x=Date, y=average,color = "2019" ),linetype="dotted",lwd = 1)
p3<-p3+geom_line(data=data_beef %>% filter(Year==2018), aes(x=Date, y=average, color = "2018"),linetype="dashed",lwd = 1)
p3<-p3+scale_color_discrete(name = "")
p3<-p3+ scale_x_date(labels = date_format("%b"),limits = as.Date(c("2020-01-27","2020-12-01")),date_breaks = "months")
p3<-p3+theme(legend.position = c(.97, .97),legend.justification = c("right", "top"),legend.key.height= unit(0.25, 'cm'),
        legend.key.width= unit(0.50, 'cm'),legend.title =element_blank(),legend.background=element_blank())


p3<-p3+theme(plot.title = element_text(face = "plain",hjust = 0.5), panel.grid.minor.x = element_blank(),axis.ticks = element_blank(),axis.title.x = element_blank())
p3<-p3+labs(y = "Average price")
p3<-p3+font("ylab")



library("cowplot")
pall<-ggdraw() +
  draw_plot(p1, x = 0, y = 0.58, width = 0.985, height = 0.42) +
  draw_plot(p2, x = 0, y = .29, width = 0.985, height = .29) +
  draw_plot(p3, x = 0, y = 0, width = 0.985, height = 0.29)

pall<-pall+ geom_text(data = data.frame(x = 0.04, y = 0.87, label = c("10^5")),aes(x, y, label = c("10^5")),
     size = 3.2,
    inherit.aes = FALSE)
#Scale for p2
pall<-pall+ geom_text(data = data.frame(x = 0.05, y = 0.59, label = paste('10^3')),aes(x, y, label = label),
     size = 3.2,
    inherit.aes = FALSE)
#right label for p1
pall<-pall+ geom_text(data = data.frame(x = 0.99, y = 0.74),aes(x, y, label = "COVID"),
     size = 4.0, angle=270,
    inherit.aes = FALSE)
#right label for p2
pall<-pall+ geom_text(data = data.frame(x = 0.99, y = 0.46, label = "M lb per day"),aes(x, y, label = label),
     size = 4.0, angle=270,  inherit.aes = FALSE)
#right label for p3
pall<-pall+ geom_text(data = data.frame(x = 0.99, y = 0.17, label = "Beef products"),aes(x, y, label = label),
     size = 4.0, angle=270,    inherit.aes = FALSE)

pall

```

The above chart answers the question mentioned at the start of this home work.

The above chart shows that increase of COVID-19 cases was significant around mid-march. The outbreak of COVID-19 in the meat processing plants was also started around mid-march as shown by the orange point marker in the chart and the cases among the meat plant workers were increased significantly by the start of April as reported by the agencies such as [Food & Environment Reporting Network](https://thefern.org/2020/05/charting-the-spread-of-covid-19-in-the-food-system/) and [CDC](https://www.cdc.gov/mmwr/volumes/69/wr/mm6927e2.htm). The increase in COVID-19 cases among the workers caused the meat plants to [shut down](https://www.meatpoultry.com/articles/22993-covid-19-meat-plant-map) starting from early April as shown by the red point markers. Each red point indicates the shutdown of a meat processing plant reported by the [Meatpoultry](https://www.meatpoultry.com/articles/22993-covid-19-meat-plant-map) website.  This has affected meat production as shown by a significant drop in production to around 1800 million pounds of meat by the end of April as compared to previous years. The meat production at the start of May was significantly low compared to the previous two years. 

The [Executive Order](https://www.usda.gov/media/press-releases/2020/04/28/usda-implement-president-trumps-executive-order-meat-and-poultry) by United States [President](https://www.washingtonpost.com/) to keep meat and poultry processing facilities to open during the COVID-19 national emergency under [Defense Production Act of 1950](https://www.whitehouse.gov/presidential-actions/executive-order-delegating-authority-dpa-respect-food-supply-chain-resources-national-emergency-caused-outbreak-covid-19/)  have forced the meat processing plants to restart meat production. The meat processing plants reopening started shortly after the Executive Order as shown by the light green point markers. The meat production rate has increased as the meat processing plants reopened and reached a similar production rate as the previous year by the start of July. The shortage of meat production during the April-May months has caused a significant increase in the beef product prices by the start of July. The price dropped after July as the meat production presumed around the start of May after initial shutdowns.
  
To design the chart, the below steps were followed:

1) First individual plots for COVID-19, meat production rates, and beef price data were generated.
2) The COVID data was plotted using  the ```ggplot()``` and ```geom_line()``` functions. In this plot, daily new COVID-19 cases were plotted using a blue line marker between the end of January to December. The heading for the chart is generated using the ```title```, ```subtitle``` features of ```labs()``` function. The legend for this plot was removed using the ```theme()``` function. The COVID-19 data plot was saved to the variable, obtained.
3) Next meat production data is plotted using the ```ggplot()``` and ```geom_line()``` functions. The approximate start of the COVID-19 outbreak in the meat processing plants is plotted using the ```geom_point()``` function with orange color point marker based on the data ontained from the [CDC](https://www.cdc.gov/mmwr/volumes/69/wr/mm6931a2.htm). The shutdowns of different meat processing plants were plotted using the ```geom_point()``` function based on the timeline provided by the [Meatpoultry](https://www.meatpoultry.com/articles/22993-covid-19-meat-plant-map) website. The red point markers indicate shut down of each meat production plant reported on the Meatpoultry website. To plot the red color point markers on the meat production line marker of 2020, the meat production rates on the dates of plant shutdowns were obtained using the ```approx()``` function. The Executive Order was annotated on the plot using ```geom_point()```, ```geom_text()``` and ```geom_segment``` functions with a dark green color point marker. Shortly after the executive order, the meat processing plants restarted production as shown by the light green points.  The meat production plot was saved to variable ```p2```.

4) The beef price data was plotted and save to variable, ```p3```. This plot is build using some of the features used for generating the above plots and the legend was also included in this plot. The legend is adjusted to be inside the plot using ```legend.justification``` feature in ```theme()``` function. For p1, p2, and p3 plots, the same x-axis limits were used.
5) Finally, all three plots were combined into a single chart using  the```draw_plot()``` and ```ggdraw()``` functions available under the ```cowplot``` library. Additional text was printed on the right of each plot using the ```geom_text()``` function and setting the relative position of the text.






## References

Data

* COVID-19 data: https://github.com/nytimes/covid-19-data
* Ground beef price: https://beta.bls.gov/dataViewer/view/timeseries/APU0000703112
* Ground chunk Beef: https://beta.bls.gov/dataViewer/view/timeseries/APU0000703111
* Uncooked ground beef: https://beta.bls.gov/dataViewer/view/timeseries/APU0000FC1101
* Meat production data: https://www.ers.usda.gov/data-products/livestock-meat-domestic-data/livestock-meat-domestic-data/#All%20meat%20statistics

* Meat shut down time line: https://www.meatpoultry.com/articles/22993-covid-19-meat-plant-map
* No fresh wave: https://www.wsj.com/articles/fresh-wave-of-meat-plant-shutdowns-unlikely-jbs-usa-chief-says-11601935477
* COVID outbreak affection on meat processing plant in South Dakota: https://www.cdc.gov/mmwr/volumes/69/wr/mm6931a2.htm
* COVID outbreak affection on meat processing plant in Virginia: https://www.vdh.virginia.gov/coronavirus/2020/09/01/covid-19-outbreaks-associated-with-meat-and-poultry-processing-plants-in-virginia-3/
* Production plants shut down: https://www.meatpoultry.com/articles/22993-covid-19-meat-plant-map
* Covid: https://thefern.org/2020/05/charting-the-spread-of-covid-19-in-the-food-system/


R Code

* For creating new columns using within Function: 
  https://stackoverflow.com/questions/39514174/adding-new-column-with-diff-function-when-there-is-one-less-row-in-r
* Working with Dates(mdy): https://mgimond.github.io/ES218/Week02c
* For diff function: https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/diff

* Separating single columns to multiple columns:  https://tidyr.tidyverse.org/reference/separate.html
* COnverting month name to a number: https://stackoverflow.com/questions/6549239/convert-months-mmm-to-numeric
* COmbining year and month: https://stackoverflow.com/questions/39420136/combine-separate-year-and-month-columns-into-single-date-column
* Display only month on the ggplot axis: https://stackoverflow.com/questions/16596811/display-the-x-axis-on-ggplot-as-month-only-in-r
* Changing colors in aes automatically: http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually#prepare-the-data
* Extracting year from date: https://www.datasciencemadesimple.com/get-year-from-date-in-r-2/
* Converting character to date datatype: https://stackoverflow.com/questions/4310326/convert-character-to-date-in-r
* Filtering year from date: https://blog.exploratory.io/filter-with-date-function-ce8e84be680
* Month scale (scale_x_date) on x-axis: https://www.datanovia.com/en/blog/ggplot-date-axis-customization/
* Different line types in R: https://www.datanovia.com/en/blog/line-types-in-r-the-ultimate-guide-for-r-base-plot-and-ggplot/
* Legends: https://www.r-graph-gallery.com/239-custom-layout-legend-ggplot2.html
*The ggplot linewidth: https://stackoverflow.com/questions/44478362/line-thickness-in-plot-function-in-r/44478798
* Converting the comma-separated character to integer:https://stackoverflow.com/questions/1523126/how-to-read-data-when-some-numbers-contain-commas-as-thousand-separator
* Looping over the dataframe rows: https://campus.datacamp.com/courses/intermediate-r-for-finance/loops-3?ex=10
* Multiple figures using ggplot: https://www.datanovia.com/en/lessons/combine-multiple-ggplots-into-a-figure/
* Remove legend name in ggplot: https://stackoverflow.com/questions/14771546/remove-legend-title-in-ggplot
* Approximation function: https://astrostatistics.psu.edu/su07/R/html/stats/html/approxfun.html
* Drawing arrows: https://stackoverflow.com/questions/38008863/how-to-draw-a-nice-arrow-in-ggplot2
*Change legend size: https://www.statology.org/ggplot2-legend-size/
